Transcriptomic Analysis for Prognostic Value in Head and Neck Squamous Cell Carcinoma

### 2021-07-01 main.tex
Updates from Overleaf

%%%%%%%%%%
% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The survival analysis of the Cancer Genome Atlas (TCGA) dataset is a well-known method to discover the gene expression-based prognostic biomarkers of head and neck squamous cell carcinoma (HNSCC). It is necessary to determine a cutoff point by the patients' dichotomization for the continuous gene expression in survival analysis. There is some optimization software for cutoff determination. However, the software's predetermined cutoffs are usually set at the median or quantiles of RNA sequencing value to perform the survival analyses. There are few clinicopathological features available on their pre-processed data sets.
We developed a workflow that includes data retrieving and pre-processing, feature selection, sliding-window cutoff mining, Kaplan-Meier survival analysis, Cox proportional hazard modeling, and biomarker discovery.
The independent HNSCC dataset has been used to validate this discovery.
Using this workflow on the TCGA HNSCC cohort, we scanned human protein-coding genes. After adjustment with confounders, we found that the clinical tumor stage and the surgical margin involvement are independent risk factors in patient survival. According to the resulting tables with Bonferroni adjusted P-value under the optimal cutoff and the hazard ratio, three biomarker candidates, CAMK2N1, IL19, and FCGBP, are significantly associated with the patients' overall survival.
We suggest the transcriptomic analysis could help for biomarker discovery.}

% Keywords
\keyword{Head and Neck Squamous Cell Carcinoma (HNSCC);
the Cancer Genome Atlas (TCGA); Transcriptomic Analysis; Survival Analysis; Optimal Cutoff; Effect Size; CAMK2N1; IL19; FCGBP; Mindfulness Meditation}

@@

%% TP53 gene => p53 protein; mutation of TP53
Accumulative knowledge showed that some biomarkers have prognostic significance in \acrshort{hnscc}. For example, node-negative \acrshort{hnscc} patients with p53 overexpression were found to hold lower survival\cite{DeVicente2004}.
Overexpression of \acrfull{hif}-1 alpha\cite{Aebersold2001} or Ki-67\cite{Couture2002} was found to be correlated with poor response to radiotherapy of \acrshort{hnscc}. The \acrfull{egfr}\cite{O-Charoenrat2000}\cite{Bentzen2005} and \acrfull{mmp}\cite{Harrington2017} were found to be overexpressed to promote invasion and metastasis of \acrshort{hnscc}.
From 2000 to 2006, the first anti-\acrshort{egfr} antibody-drug (cetuximab) has been developed and combined with radiotherapy, known as bio-\acrshort{rt}, to increase survival of unresectable locoregionally advanced disease\cite{Bonner2006a}.
The systemic therapy of cetuximab plus platinum-fluorouracil chemotherapy (EXTREME regimen) has improved overall survival when given as first-line treatment in patients with recurrent or metastatic \acrshort{hnscc}\cite{Vermorken2008}\cite{Rivera2009}. It has been approved by the \acrshort{us} \acrfull{fda} since 2008. In advance, the bio-\acrshort{rt} could be proceeded with \acrfull{tpf} induction chemotherapy to overcome  radio-resistance of \acrshort{hnscc}\cite{Blanchard2013}.

However, Rampias and his colleagues\cite{Rampias2014} suggested that \acrfull{hras} mutations could mediate cetuximab resistance in systemic therapy of \acrshort{hnscc} via the \acrshort{egfr}/\acrfull{ras}/\acrfull{erk} signaling pathway.
After that, the \acrshort{egfr} \acrfull{tki} was introduced to help cetuximab in 2018. The anti-tumor activity was observed in a phase 1 trial for \acrshort{hnscc} patients using cetuximab and afatinib, a \acrshort{tki} of \acrshort{egfr}, \acrfull{her}2, and \acrshort{her}4\cite{Gazzah2018}. Other \acrshort{egfr} \acrshort{tki}s, such as gefitinib, erlotinib, osimertinib, were also developed to treat advanced \acrshort{hnscc}.
Although 90\% of \acrshort{hnscc} has overexpression of \acrshort{egfr}, cetuximab has only 10\% to 20\% response rate on those patients. So far, cetuximab is still the only drug of choice with proven efficacy, which has targeted the selected \acrshort{hnscc} patients\cite{Taberna2019}.

@@

According to our proteomic study between 2010 to 2017, \acrfull{tmsb4x} was reported to be related to tumor growth and metastasis of \acrshort{hnscc}\cite{Chi2017}. It was also reported by the subsequent investigations that \acrshort{tmsb4x} has engaged in tumor aggressiveness through \acrfull{emt} on pancreatic\cite{Zhang2008}, gastric\cite{Ryu2012}, colorectal\cite{Gemoll2015}, lung\cite{Huang2016}, ovarian\cite{Chu2019}, and melanoma\cite{Makowiecka2019} cancers. Thus, it might be suggested that \acrshort{tmsb4x}  is a candidate for tumor type-agnostic therapy\cite{Yan2018} as a common biomarker crossing several types of cancer.
%TMSB4X proteins regulate intracellular signal transduction and has been proved to overexpress in various cancers, including colorectal, lung, gastric, pancreatic, and squamous cell cancers.\cite{Chi2017}
% https://doi.org/10.1038/s41598-017-09539-w

\begin{MyColorPar}{red}
The Cancer Genome Atlas (\acrshort{tcga}) profiled 528 \acrshort{hnscc} clinical and genomic data, which has been standardized and available at a unified data portal, \acrfull{gdc} of the \acrfull{nci}.
%\cite{SeanDavis;MartinMorgan2020}
\end{MyColorPar} % red; the rest moved into materials and methods

% to summarize the advantages of its use.
The highlights of the advantages of applying the TCGA data for cancer biomarker identification include:
\begin{outline}
\1  To the best of our knowledge, the TCGA database is the largest collection (in both cancer types and cohort size) of comprehensive genomics with survival data available in the field of cancer research so far.  
There are many physical and social features of patients available for survival modeling. % ($X_1 ... X_n$)
The whole-genome sequencing data has been harmonized across all GDACs.  Many databases adopt the essential demographic data from TCGA since it has comprehensive clinical features, such as exposure to alcohol, asbestos, radioactive radon, tobacco smoking, or cigarettes.
\1  TCGA has a remarkable advantage for computational and life scientists who study cancer since useful web-based tools and API are ready to analyze and visualize TCGA data. It might be getting help quickly from the research community for trouble-shooting.
\1  Many achievements of diagnoses, treatments, and prevention with the TCGA data have already been published and keep growing\cite{Tomczak2015}.\\[0.5cm]
\end{outline}

\begin{MyColorPar}{red}
%\subsubsection{From TCGA to expression biomarkers} % from Answer 2-6 minor2
Usually, researchers developed an in-house workflow of gene expression analysis to find HNSCC biomarkers.
They have several genes that are differentially expressed between tumor and normal samples.
It would be helpful to show that alteration in gene expression correlates with phenotypes of HNSCC

% Rstudio\cite{Loraine2015a}
Some researchers\cite{Loraine2015a}\cite{Tonella2017a}\cite{Zhao2018}\cite{Li2018a}\cite{Huang2019}\cite{Shen2019}\cite{Schmitt2019}\cite{Xu2021a} tried to find \acrfull{degs} of the HNSCC samples at both genotypic and phenotypic levels (without survival information) for biomarker discovery.
Gene expression data was downloaded from the TCGA or Gene Expression Omnibus (GEO) database (e.x. GSE117973\cite{Schmitt2019}, HIPO-HNC cohort has n = 87).
% IRB: Patient samples were obtained under the protocol S-206/2011, approved by the Ethics Committee of Heidelberg University, with written informed consent from all participants.
%e.x. GSE30784: 167 HNSCC samples and 45 normal controls). 
They employed the Database for Annotation, Visualization, and Integrated Discovery (DAVID, available at https://david.ncifcrf.gov/) to obtain information for Gene Ontology (GO), including biological processes, the cellular component, and molecular function. 
The Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway analysis was also used to annotate the potential functions.
%Subsequently, comprehensive bioinformatics analysis incorporating gene ontology (GO), Kyoto Encyclopedia of Genes and Genomes (KEGG),
The protein-protein interaction (PPI) network was conducted by STRING (available at https://string-db.org).
The pathway enrichment analysis of \acrshort{degs} was also performed by DAVID, String, or Cytoscape software programs\cite{Huang2019}\cite{Shen2019}.
Li\cite{Li2018a} and his colleagues made an R package (GDCRNATool) for the implementation of those workflows on gene expression analyses of the TCGA.
%The DEmRNAs were enrolled in a protein-protein interaction (PPI) network through the STRING database (https://string-db.org/) with a confidence score $> 0.9$, and the PPI network was visualized in Cytoscape (Version 3.7.1) software. Moreover, genes with degree $>= 25$ were selected as hub genes. Subsequently, module analysis (16) of the PPI network was performed using the Molecular Complex Detection (MCODE) tool of Cytoscape software, and GO and KEGG analysis of the modules was carried out using the DAVID database.
%a validation set and
%Then, Kaplan–Meier survival curves along with a log-rank test were applied to select candidates\cite{Xu2021a}.
Xu and his colleagues\cite{Xu2021a} also used Kaplan–Meier survival curves along with a log-rank test or univariate Cox regression to screen prognostic factors (\textit{P} value \begin{equation}< 0.05 \end{equation}).
The gene expression was used to select the preliminary biomarkers with a significant impact on patients' overall survival (\textit{P} value $< 0.01$).
%By integrated bioinformatics analysis, including protein_protein interaction (PPI) network, Gene Ontology (GO) enrichment, and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analyses. The PPI network was established using the Search Tool for the Retrieval of Interacting Genes (STRING) and Cytoscape software.
The candidate genes were validated via the Gene Expression Profiling Interactive Analysis tool (GEPIA1/2 database using TCGA datasets, developed by Zefang Tang and his colleagues\cite{Tang2017a}\cite{Tang2019}
%, datasets from projects of the TCGA and the Genotype-Tissue Expression, GTEx
), GEPIA2021\cite{Li2021} (available at http://\\gepia2021.cancer-pku.cn/), or the Human Protein Atlas (HPA) website.
% GTEx https://gtexportal.org/home/; non-disease 54 tissues/organs
% GEPIA1/2 contact us: tangzefang@pku.edu.cn
Finally, the gene expression profile from the GEO and HNSCC cell lines and tissues were used to verify these biomarkers.\\
\end{MyColorPar} % red

@@

The other investigators should get the rationale or revelation of the genes of interest on a specific cancer type. They should upload those genes manually onto bioinformatics tools, such as SurvExpress (http://bioinformatica.mty.itesm.mx:8080/Biomatec/ \newline
SurvivaX.jsp\cite{Aguirre-Gamboa2013}, which has been lost since Oct/2019 and currently out of funds), and analyze the cohort (e.x. \acrshort{tcga}). After downloading the survival results, they could curate plots and tables carefully.
It is not possible to scan the whole human protein-coding genome in this way.
The web-based tools might set a cutpoint at the median, 1/4 quantile, or 3/4 quantile for subsequent analyses. There are several visualization software or R packages which deal with cutoff determination\cite{Abel1984}, %CRITLEVEL
such as Prognoscan\cite{Mizuno2009a}, Cutoff Finder\cite{Budczies2012}, Findcut\cite{Chang2017a}, Human protein atlas\cite{Uhlen2017}, OptimalCutpoints\cite{Cristina2019},  cutpointr (available at https://github.com/thie1e/cutpointr), and cutoffR (available at https://cran.r-project.org/web/
packages/cutoffR). 
However, non of them could perform jointly predictive of survival times and cutoff selection within subgroups of gene expression.
%  %combine the scanning of the protein-coding genes and cutoff optimization programmatically.

@@

First, although \acrshort{tcga} genomics data has been harmonized, 
%two things were found during our investigating and data cleaning:
the pre-processing of TCGA RNA-Seq is as follows:

\begin{outline}
\1 HNSCC samples without complete clinical information have been removed;\\
% or 50\%
\1 Null expressed genes in more than 30\% of the HNSCC samples have been excluded;\\

\1 Updated number of protein-coding genes in the TCGA HNSCC is 20500.\\
After investigation of mRNA expression dataset obtained through NCI's Firehose API, we found that gene ID (9906), gene ID (728661), and their expression values were saved together under the entity of gene symbol "SLC35E2".
The expression file of SLC35E2 has almost double in size than those of SLC35E1 or SLC35E3.
% http://gepia2.cancer-pku.cn/#survival is SLC35E2; 證明 TCGA has error
% [line 1549]
%HNSCC.mRNA.Exp.SLC35E2.Fire.Rda
%[line 1654]
%for gene correction (20499 to 20500) from SLC35E2 to 2 genes:
%NCBI Reference Sequences (RefSeq) National Center for Biotechnology Information Reference Sequence (NCBI) RefSeq Release 38 (November 7, 2009) 
According to the Human Gene Database (available at https://www.genecards.org/Search\\/Keyword?queryString=SLC35E2), SLC35E2A (Gene ID: 9906) and\\ SLC35E2B (Gene ID: 728661) should be the correct entities for TCGA HNSCC dataset.
% https://www.ncbi.nlm.nih.gov/gene/?term=(SLC35E2)+AND+%22Homo+sapiens%22
% keyword: (SLC35E2) AND "Homo sapiens" 
SLC35E2 is the previous symbol of SLC35E2A (reference at https://www.genenames.org/data/gene-symbol-report/\#!\\/hgnc\_id/HGNC:20863).
Thus, we made reassignment of the expression value of SLC35E2A and SLC35E2B and updated the number of protein-coding genes in the TCGA HNSCC from 20499 to 20500.
% updated on 18-May-2021
\end{outline}

@@

% dissection
%ZSWIM2\_archive.Rda % 其實最困難的在於 unknown error (ZSWIM2) 
Third, we have analyzed the error log of the cutoff finding and Cox modeling, 
the results show that several technical problems could halt the program running.
Those include:
\begin{outline}

\1 32.2\% event has "one group" issue in confusion matrix of Chi-square test in Cox regression
\1 21.05\% error occurs by "one group" issue in log-rank test (survdiff or survdiff.fit function in R package "survival") in Kaplan-Meier estimate.
%1) ok [1] "[ 4 / 20083 ] Generating median-cut KM plot of A1BG"
%Error in survdiff.fit(y, groups, strata.keep, rho) : at A1CF
% error at 16115 gene ""; why? 518 values all are -0.04203 at T
%  There is only 1 group 
%This is because of skewed distribution of the expression value. 
% 67/518 =  13% cases >1, while 450 (87%) cases = -0.30949; so median is -0.30949; 
% all cases belong to >= median => 改為 > median 即可
% 2) % survfit 16114 has data set has no non-missing observations: NA's 113/518 cases
% median = NaN :-); 
% solved by cutoff <- median(gset_OS_mRNA2[, i], na.rm = TRUE)
% and NaN imputation as median
\1 0.78\% has unknown reason (so those 159 genes has been excluded in our workflow)
\end{outline}
, which could not be detected prior to the program running.
It might be due to skewed distribution of the expression value or even random error in 20500 RNA-Seq data. % heterogeneity

@@

Our approach describes a workflow implemented in the R script, which runs on the Rstudio server.
Its function includes data retrieving and pre-processing, feature selection, sliding-window cutoff mining, Kaplan-Meier survival analysis, Cox proportional hazard modeling, and biomarker discovery.
The independent HNSCC dataset has been used to validate this discovery.
The workflow, shown in Figure \ref{fig:figure1}, scanned 20,500 human protein-coding genes of the \acrshort{tcga} \acrshort{hnscc} cohort to yield a model with biomarker estimate.

@@

%\subsubsection{Figure 1:h}
\begin{figure}[hbt!]
\centering
%\widefigure
\includegraphics[width=14cm]{Figure_1_manuscript_workflow} % .PDF is better than .png
%, height=8cm
%\caption % , step 1 (\textcolor{blue}{blue line}: main procedure) and step 2 (\textcolor{orange}{orange line}: analysis export).
%Step 3 (purple line: dealing with surgical margin).
\bcaption{A workflow of \acrshort{hnscc} biomarker discovery}
{The workflow includes data retrieval from TCGA GDC data portal, data process with merging and cleaning, then performing the survival analyses (within \textcolor{yellow}{yellow} square). The Cutoff engine (in R script: cutofFinder\_func.HNSCC.R, a serial cut for grouping patients with \textcolor{asparagus}{low} or \textcolor{red}{high} expression of a specific gene, to yield a collection of \textit{P} values; please see Materials and Methods section for details) might calculate all possible Kaplan-Meier \textit{P} values (corrected by \acrlong{fdr}, FDR, method) to find the optimal cutoff value of gene expression for subsequent Cox modeling.
The candidate selection performs (1) dissecting and selection of candidate genes by further Bonferroni adjusted \textit{P} values as well as a hazard ratio of Cox model, based on the results from the survival analyses;
(2) survival analyses of the other HNSCC dataset (GSE6585) using Kaplan-Meier estimate (KM, with FDR correction) and Cox modeling (CoxHR).\\
The selected genes have been validated by third HNSCC cohort (GSE2837).
(HNSCC: head and neck squamous cell carcinoma; TCGA: the Cancer Genome Atlas; RNA-Seq: RNA sequencing; GDC: Genomic Data Commons.)}
\label{fig:figure1}
% Description:1) FDR correction of Kaplan-Meier \textit{P} values during Cutoff finding; and 2) Bonferroni correction of Kaplan-Meier \textit{P} values after Cox modeling for candidate selection.

@@

@@

% end of result
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Example of a figure that spans the whole page width (the commands \widefigure and \begin{paracol}{2}, \linenumbers, and\switchcolumn need to be present). The same concept works for tables, too.
%\begin{figure}[H]	
%\widefigure
%\includegraphics[width=15 cm]{Definitions/logo-mdpi}
%\caption{This is a wide figure.\label{fig2}}
%\end{figure}

\begin{paracol}{2}
\linenumbers
\switchcolumn

>>>>

%%%%%%%%%%
% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The survival analysis of the Cancer Genome Atlas (TCGA) dataset is a well-known method to discover the gene expression-based prognostic biomarkers of head and neck squamous cell carcinoma (HNSCC). It is necessary to determine a cutoff point by the patients' dichotomization for the continuous gene expression in survival analysis. There is some optimization software for cutoff determination. However, the software's predetermined cutoffs are usually set at the median or quantiles of RNA sequencing value to perform the survival analyses. There are few clinicopathological features available on their pre-processed data sets.
We developed a workflow that includes data retrieving and pre-processing, feature selection, sliding-window cutoff mining, Kaplan-Meier survival analysis, Cox proportional hazard modeling, and biomarker discovery.
Using this workflow on the TCGA HNSCC cohort, we scanned human protein-coding genes. After adjustment with confounders, we found that the clinical tumor stage and the surgical margin involvement are independent risk factors in patient survival. According to the resulting tables with Bonferroni adjusted P-value under the optimal cutoff and the hazard ratio, three biomarker candidates, CAMK2N1, IL19, and FCGBP, are significantly associated with the patients' overall survival. The other independent HNSCC data were also used to validate this discovery.
We suggest the transcriptomic analysis could help for biomarker discovery.}

% Keywords
\keyword{Head and Neck Squamous Cell Carcinoma (HNSCC);
the Cancer Genome Atlas (TCGA); Transcriptomic Analysis; Survival Analysis; Optimal Cutoff; Effect Size; \acrfull{CAMK2N1}; \acrfull{IL19}; \acrfull{FCGBP}; Mindfulness Meditation}

@@

%% TP53 gene => p53 protein; mutation of TP53
Accumulative knowledge shows that some biomarkers have prognostic significance in \acrshort{hnscc}. For example, node-negative \acrshort{hnscc} patients with p53 overexpression were found to hold lower survival\cite{DeVicente2004}.
Overexpression of \acrfull{hif}-1 alpha\cite{Aebersold2001} or Ki-67\cite{Couture2002} was found to be correlated with poor response to radiotherapy of \acrshort{hnscc}. The \acrfull{egfr}\cite{O-Charoenrat2000}\cite{Bentzen2005} and \acrfull{mmp}\cite{Harrington2017} were found to be overexpressed to promote invasion and metastasis of \acrshort{hnscc}.
From 2000 to 2006, the first anti-\acrshort{egfr} antibody-drug (cetuximab) has been developed and combined with radiotherapy, known as bio-\acrshort{rt}, to increase survival of unresectable locoregionally advanced disease\cite{Bonner2006a}.
The systemic therapy of cetuximab plus platinum-fluorouracil chemotherapy (EXTREME regimen) improves overall survival when given as first-line treatment in patients with recurrent or metastatic \acrshort{hnscc}\cite{Vermorken2008}\cite{Rivera2009}. It has been approved by the \acrshort{us} \acrfull{fda} since 2008. In advance, the bio-\acrshort{rt} could be proceeded with \acrfull{tpf} induction chemotherapy to overcome  radio-resistance of \acrshort{hnscc}\cite{Blanchard2013}.

However, Rampias and his colleagues\cite{Rampias2014} suggested that \acrfull{hras} mutations could mediate cetuximab resistance in systemic therapy of \acrshort{hnscc} via the \acrshort{egfr}/\acrfull{ras}/\acrfull{erk} signaling pathway.
After that, the \acrshort{egfr} \acrfull{tki} was introduced to help cetuximab in 2018. The anti-tumor activity was observed in a phase 1 trial for \acrshort{hnscc} patients using cetuximab and afatinib, a \acrshort{tki} of \acrshort{egfr}, \acrfull{her}2, and \acrshort{her}4\cite{Gazzah2018}. Other \acrshort{egfr} \acrshort{tki}s, such as gefitinib, erlotinib, osimertinib, were also developed to treat advanced \acrshort{hnscc}.
Although 90\% of \acrshort{hnscc} has overexpression of \acrshort{egfr}, cetuximab has only 10\% to 20\% response rate on those patients. Before the year 2019, cetuximab is still the only drug of choice with proven efficacy, which has targeted the selected \acrshort{hnscc} patients\cite{Taberna2019}.

@@

According to our previous proteomic study from 2010 to 2017, \acrfull{tmsb4x} was reported to be related to tumor growth and metastasis of \acrshort{hnscc}\cite{Chi2017}. It was also reported by the subsequent investigations that \acrshort{tmsb4x} has engaged in tumor aggressiveness through \acrfull{emt} on pancreatic\cite{Zhang2008}, gastric\cite{Ryu2012}, colorectal\cite{Gemoll2015}, lung\cite{Huang2016}, ovarian\cite{Chu2019}, and melanoma\cite{Makowiecka2019} cancers. Thus, it might be suggested that \acrshort{tmsb4x}  is a candidate for tumor type-agnostic therapy\cite{Yan2018} as a common biomarker crossing several types of cancer.
%TMSB4X proteins regulate intracellular signal transduction and has been proved to overexpress in various cancers, including colorectal, lung, gastric, pancreatic, and squamous cell cancers.\cite{Chi2017}
% https://doi.org/10.1038/s41598-017-09539-w

%\begin{MyColorPar}{red}
The Cancer Genome Atlas (\acrshort{tcga}) profiled \acrshort{hnscc} (528 participants) clinical and genomic data, which were standardized and are available at a unified data portal, \acrfull{gdc} of the \acrfull{nci}.
%\cite{SeanDavis;MartinMorgan2020}
%\end{MyColorPar} % red; the rest moved into materials and methods
% to summarize the advantages of its use.
The highlight of the advantages of applying the TCGA data for cancer biomarker identification includes:
\begin{outline}
\1  To the best of our knowledge, the TCGA database is the largest collection (in both cancer types and cohort size) of comprehensive genomics with survival data available in the field of cancer research so far.  
There are many physical and social features of patients available for survival modeling. % ($X_1 ... X_n$)
The whole-genome sequencing data were harmonized across all \acrfull{gdac}s.  Many databases adopt the essential demographic data from TCGA since it has comprehensive clinical features, such as exposure to alcohol, asbestos, radioactive radon, tobacco smoking, or cigarettes.
\1  TCGA has a remarkable advantage for computational and life scientists who study cancer since useful web-based tools and API are ready to analyze and visualize TCGA data. It might be getting help quickly from the research community for trouble-shooting.
\1  Many achievements of diagnoses, treatments, and prevention with the TCGA data were already published and keep growing\cite{Tomczak2015}.\\[0.1cm]
\end{outline}

\begin{MyColorPar}{red}
%\subsubsection{From TCGA to expression biomarkers} % from Answer 2-6 minor2
Usually, researchers developed an in-house workflow of gene expression analysis of TCGA data to find HNSCC biomarkers.
They have several genes that are differentially expressed between tumor and normal samples.
It would be helpful to show that alteration in gene expression correlates with phenotypes of HNSCC
% Rstudio\cite{Loraine2015a}
Some researchers\cite{Loraine2015a}\cite{Tonella2017a}\cite{Zhao2018}\cite{Li2018a}\cite{Huang2019}\cite{Shen2019}\cite{Schmitt2019}\cite{Xu2021a} tried to find \acrfull{degs} of the HNSCC samples at both genotypic and phenotypic levels (without survival information) for biomarker discovery.
Gene expression data was downloaded from the TCGA or Gene Expression Omnibus (GEO) database (e.g., GSE117973\cite{Schmitt2019}, HIPO-HNC cohort has n = 87).
% IRB: Patient samples were obtained under the protocol S-206/2011, approved by the Ethics Committee of Heidelberg University, with written informed consent from all participants.
%e.g., GSE30784: 167 HNSCC samples and 45 normal controls). 
They employed the Database for Annotation, Visualization, and Integrated Discovery (DAVID, available at https://david.ncifcrf.gov/) to obtain information for Gene Ontology (GO), including biological processes, the cellular component, and molecular function. 
The Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway analysis was also used to annotate the potential functions.
%Subsequently, comprehensive bioinformatics analysis incorporating gene ontology (GO), Kyoto Encyclopedia of Genes and Genomes (KEGG),
The protein-protein interaction (PPI) network was conducted by STRING (available at https://string-db.org).
The pathway enrichment analysis of \acrshort{degs} was also performed by DAVID, String, or Cytoscape software programs\cite{Huang2019}\cite{Shen2019}.
Li\cite{Li2018a} and his colleagues made an R package (GDCRNATool) for the implementation of those workflows on gene expression analyses of the TCGA.
%The DEmRNAs were enrolled in a protein-protein interaction (PPI) network through the STRING database (https://string-db.org/) with a confidence score $> 0.9$, and the PPI network was visualized in Cytoscape (Version 3.7.1) software. Moreover, genes with degree $>= 25$ were selected as hub genes. Subsequently, module analysis (16) of the PPI network was performed using the Molecular Complex Detection (MCODE) tool of Cytoscape software, and GO and KEGG analysis of the modules was carried out using the DAVID database.
%a validation set and
%Then, Kaplan–Meier survival curves along with a log-rank test were applied to select candidates\cite{Xu2021a}.
Xu and his colleagues\cite{Xu2021a} also used Kaplan–Meier survival curves with a log-rank test, and univariate Cox regression to screen prognostic factors.
The gene expression was used to select the preliminary biomarkers with a significant impact on patients' overall survival (\textit{P} value $< 0.01$).
%By integrated bioinformatics analysis, including protein_protein interaction (PPI) network, Gene Ontology (GO) enrichment, and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analyses. The PPI network was established using the Search Tool for the Retrieval of Interacting Genes (STRING) and Cytoscape software.
The candidate genes were validated via the Gene Expression Profiling Interactive Analysis tool (GEPIA1/2 database using TCGA datasets, developed by Zefang Tang and his colleagues\cite{Tang2017a}\cite{Tang2019}
%, datasets from projects of the TCGA and the Genotype-Tissue Expression, GTEx
), GEPIA2021\cite{Li2021} (available at http://gepia2021.cancer-pku.cn/), or the Human Protein Atlas (HPA) website.
% GTEx https://gtexportal.org/home/; non-disease 54 tissues/organs
% GEPIA1/2 contact us: tangzefang@pku.edu.cn
Finally, the gene expression profile from the GEO and HNSCC cell lines and tissues were used to verify these biomarkers.\\
\end{MyColorPar} % red

@@

The other investigators should get the rationale or revelation of the genes of interest on a specific cancer type. They should upload those genes manually onto bioinformatics tools, such as SurvExpress (http://bioinformatica.mty.itesm.mx:8080/Biomatec/ \newline
SurvivaX.jsp\cite{Aguirre-Gamboa2013}, which has been lost since Oct/2019 and currently out of funds), and analyze the cohort (e.g., \acrshort{tcga}). After downloading the survival results, they could curate plots and tables carefully.
It is not possible to scan the whole human protein-coding genome in this way.
The web-based tools might set a cutoff at the median, 1/4 quantile, or 3/4 quantile for subsequent analyses. There are several visualization software or R packages which deal with cutoff determination\cite{Abel1984}, %CRITLEVEL
such as Prognoscan\cite{Mizuno2009a}, Cutoff Finder\cite{Budczies2012}, Findcut\cite{Chang2017a}, Human protein atlas\cite{Uhlen2017}, OptimalCutpoints\cite{Cristina2019},  cutpointr (available at https://github.com/thie1e/cutpointr), and cutoffR (available at https://cran.r-project.org/web/
packages/cutoffR). 
However, non of them could perform jointly predictive of survival times and cutoff selection within subgroups of gene expression.
%  %combine the scanning of the protein-coding genes and cutoff optimization programmatically.

@@

First, although \acrshort{tcga} genomics data were harmonized, 
% three things were found during our investigating and data cleaning:
the pre-processing of TCGA RNA-Seq in our workflow has been done as follows: 
\begin{outline}
\1 HNSCC samples without complete clinical information have been removed;
\1 Null expressed genes in more than 30\% of the HNSCC samples have been excluded; % or 50\%
\1 Updated number of protein-coding genes in the TCGA HNSCC is 20500.
\end{outline}
After investigation of mRNA expression dataset obtained through NCI's Firehose API, we found that expression value of two genes (gene ID: 9906 and gene ID: 728661) was saved together under the entity of gene symbol "SLC35E2".
The expression file of SLC35E2 has almost double in size than those of SLC35E1 or SLC35E3.
% http://gepia2.cancer-pku.cn/#survival is SLC35E2; 證明 TCGA has error
% [line 1549]
%HNSCC.mRNA.Exp.SLC35E2.Fire.Rda
%[line 1654]
%for gene correction (20499 to 20500) from SLC35E2 to 2 genes:
%NCBI Reference Sequences (RefSeq) National Center for Biotechnology Information Reference Sequence (NCBI) RefSeq Release 38 (November 7, 2009) 
According to the Human Gene Database (available at https://www.genecards.org/Search/Keyword?\\queryString=SLC35E2), SLC35E2A (Gene ID: 9906) and SLC35E2B (Gene ID: 728661) should be the correct entities for the TCGA HNSCC dataset.
% https://www.ncbi.nlm.nih.gov/gene/?term=(SLC35E2)+AND+%22Homo+sapiens%22
% keyword: (SLC35E2) AND "Homo sapiens" 
SLC35E2 is the previous symbol of SLC35E2A (reference at https://www.genenames.org/data/gene-symbol-report/\#!/hgnc\_id/HGNC:20863).
Thus, we made reassignment of the expression value of SLC35E2A and SLC35E2B and updated the number of protein-coding genes in this TCGA HNSCC dataset from 20499 to 20500.
% updated on 18-May-2021
%\end{outline}

@@

% dissection
%ZSWIM2\_archive.Rda % 其實最困難的在於 unknown error (ZSWIM2) 
Third, we analyzed the error log during the cutoff finding and Cox modeling, 
the result shows that program running could be halted by several technical problems.
These include:
\begin{outline}

\1 32.2\% event has "one group" issue in confusion matrix of Chi-square test in Cox regression;
\1 21.05\% error occurs by "one group" issue in log-rank test (survdiff or survdiff.fit function in R package "survival") in Kaplan-Meier estimate;
%1) ok [1] "[ 4 / 20083 ] Generating median-cut KM plot of A1BG"
%Error in survdiff.fit(y, groups, strata.keep, rho) : at A1CF
% error at 16115 gene ""; why? 518 values all are -0.04203 at T
%  There is only 1 group 
%This is because of skewed distribution of the expression value. 
% 67/518 =  13% cases >1, while 450 (87%) cases = -0.30949; so median is -0.30949; 
% all cases belong to >= median => 改為 > median 即可
% 2) % survfit 16114 has data set has no non-missing observations: NA's 113/518 cases
% median = NaN :-); 
% solved by cutoff <- median(gset_OS_mRNA2[, i], na.rm = TRUE)
% and NaN imputation as median
\1 0.78\% has unknown reason (so those 159 genes has been excluded in our workflow).
\end{outline}
These technical problems could not be detected prior to program running.
It might be due to skewed distribution of the expression value or even random error in 20500 RNA-Seq data. % heterogeneity

@@

Our approach describes a workflow implemented in the R script, which runs on the Rstudio server.
Its function includes data retrieving and pre-processing, feature selection, sliding-window cutoff mining, Kaplan-Meier survival analysis, Cox proportional hazard modeling, and biomarker discovery.
The independent HNSCC datasets were used to validate this strategy.
The workflow, shown in Figure \ref{fig:figure1}, scanned 20,500 human protein-coding genes of the \acrshort{tcga} \acrshort{hnscc} cohort to yield a model with biomarker estimate.

@@

%\subsubsection{Figure 1:h}
\begin{figure}[hbt!]
\centering
%\widefigure
\includegraphics[width=14cm]{Figure_1_manuscript_workflow} % .PDF is better than .png
%, height=8cm
%\caption % , step 1 (\textcolor{blue}{blue line}: main procedure) and step 2 (\textcolor{orange}{orange line}: analysis export).
%Step 3 (purple line: dealing with surgical margin).
\bcaption{A workflow of \acrshort{hnscc} biomarker discovery}
{The workflow includes data retrieval from TCGA GDC data portal, data process with merging and cleaning, then performing the survival analyses (within \textcolor{yellow}{yellow} square). The Cutoff engine (in R script: cutofFinder\_func.HNSCC.R, a serial cut for grouping patients with \textcolor{asparagus}{low} or \textcolor{red}{high} expression of a specific gene, to yield a collection of \textit{P} values; please see Materials and Methods section for details) might calculate all possible Kaplan-Meier \textit{P} values (corrected by \acrlong{fdr}, FDR, method) to find the optimal cutoff value of gene expression for subsequent Cox modeling.
The candidate selection performs (1) dissecting and selection of candidate genes by further Bonferroni adjusted \textit{P} values as well as a hazard ratio of Cox model, based on the results from the survival analyses;
(2) survival analyses of the other HNSCC dataset (GSE65858) using Kaplan-Meier estimate (KM, with FDR correction) and Cox modeling (CoxHR).\\
The selected genes were validated by third HNSCC cohort (GSE2837).\\
(HNSCC: head and neck squamous cell carcinoma; TCGA: the Cancer Genome Atlas; RNA-Seq: RNA sequencing; GDC: Genomic Data Commons.)}
\label{fig:figure1}
% Description:1) FDR correction of Kaplan-Meier \textit{P} values during Cutoff finding; and 2) Bonferroni correction of Kaplan-Meier \textit{P} values after Cox modeling for candidate selection.

@@

\multicolumn{8}{|l|}{(OS: overall survival ; \textit{P}~value significant codes is denoted: red \textless{} 0.05; *** \textless{} 0.001)}                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                               \\

@@
